Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-08-27 review/Response to reviews.Rmd

<!--# need to double-check and revise -->

The difference in flowering probability between cointinuous forests and fragments is in fact significant.
We have added Table 1 that shows the model intercepts ± 84% confidence intervals, which is comparable to testing a main effect of habitat if we were able to include both habitats in the same model.

##  {#R2-7}

> Delayed effects: Since the authors fit separate models for plants in fragments and in continuous forest to allow for the crossbasis function to differ between habitats we don't know if any of the small differences reported in Fig 4c, 5c and 6c are statistically significant.
> Hence, I would recommend interpreting the results concerning the difference between fragments and CF with caution.

Because we were not able to test for a significant interaction between habitat and lagged SPEI, we did not intend to suggest that our results implied statistically significant differences in the shape of the crossbasis functions between fragments and continuous forest.
We have made changes throughout the manuscript that clarify this and remove some of the stronger declarative language about differences in the effects of SPEI between continuous forest and forest fragments.

> I would like to see at least a justification for why representing the difference figures are meaningful.

We believe the difference figures (panel c in figure 4,5, and 6) are helpful in showing combinations of SPEI and lag time where the marginal effects on vital rates are most divergent between continuous forest and fragments.
For example, at a lag of 0, wet conditions have a positive effect on survival in continuous forest and a negative effect on survival in fragments, leading to a large predicted difference in survival between habitats under those conditions.

### 2021-08-27 doc/paper.Rmd
Close #85

Understanding how landscape structure and abiotic conditions act to influence population dynamics is central to many conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
Our results support the emerging consensus that the effects of climatic extremes on demographic vital rates can be delayed for months or even years [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
We also found that the delayed responses of populations in fragments can differ substantially in magnitude, direction, and lag time from those of populations in continuous forest.
This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015; @laurancePositiveFeedbacksForest2001] are likely to be pervasive, but also far more complex than previously thought.

@@

It is important to emphasize, however, that the overall effects of SPEI on survival and growth are more severe in fragments than continuous forest (Figures \@ref(fig:surv-spei), \@ref(fig:growth-spei)). Furthermore, the magnitude of plant responses to climatic extremes is also greater in habitat fragments---extreme drought in dry seasons and extreme precipitation in during rainy seasons are most detrimental to growth and survival in fragments. While intact forest and its canopy buffer populations from climatic extremes, populations in fragments---especially near edges with high contrast matrix---likely lack this protection [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013]. We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@lauranceApparentEnvironmentalSynergism2014], that that are the causal mechanism underlying reduced plant growth and survival in forest fragments.

@@

While it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
Our results clearly indicate that they can, with habitat-specific differences in how environmental conditions influenced future vital rates.
For example, extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival in both habitat types.
However, the magnitude of these declines was far greater in forest fragments.
Similarly, the detrimental effects of extremes in SPEI on growth rates were also more pronounced in fragments.
In contrast, variation in SPEI had far stronger effects on the probability of flowering in continuous forest than fragments.
These results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.

>>>>

Understanding how landscape structure and abiotic conditions act to influence population dynamics is central to many conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
Our results support the emerging consensus that the effects of climatic extremes on demographic vital rates can be delayed for months or even years [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
We also found different magnitudes, directions, and lag times of climate effects in fragments and continuous forests.
This suggests that the hypothesized synergies between climate and fragmentation on population dynamics [@opdamClimateChangeMeets2004; @selwoodEffectsClimateChange2015; @laurancePositiveFeedbacksForest2001] may be important in this system in a way far more complex than previously thought.

@@

It is important to emphasize, however, that the range of the estimated response to SPEI for survival and growth was greater in fragments compared with continuous forest (Figures \@ref(fig:surv-spei), \@ref(fig:growth-spei)). This suggests that extremes in SPEI may be more detrimental in forest fragments compared with continuous forest. While intact forest and its canopy buffer populations from climatic extremes, populations in fragments---especially near edges with high contrast matrix---likely lack this protection [@didhamEdgeStructureDetermines1999; @ewersFragmentationImpairsMicroclimate2013]. We suggest it is these climate extremes, rather than trends in average temperature, precipitation, or SPEI [@lauranceApparentEnvironmentalSynergism2014], that that are the causal mechanism underlying the observed reduced plant growth and survival in forest fragments.

@@

While it appears that delayed effects of climate on demographic vital rates may be ubiquitous [@eversLaggedDormantSeason2021], the extent to which they vary spatially or with habitat remains an open question.
Our results suggest that they may, with habitat-specific differences in how environmental conditions influenced future vital rates.
For example, extreme values of SPEI---both positive (unusually high precipitation) and negative (drought conditions)---led to declines in the probability of individual survival in forest fragments while extreme wet conditions significantly increased survival in continuous forest.
Similarly, the marginal effects of SPEI on plant size were greater in fragments, suggesting a more pronounced effect of extreme climates in fragments.
In contrast, variation in SPEI corresponded to a greater range of flowering probabilities in continuous forest than fragments.
These results should be interpreted with some caution, however, as the relatively low number of plants in fragments that are above the threshold-size for flowering could limit the power to detect delayed effects.
